Fractional Types

### 2019-11-18 fracGC/p.tex
sec 2 kind of

\newcommand{\alt}{~|~}
\newcommand{\net}{t_\bullet}
\newcommand{\inlv}[1]{\ensuremath{\mathit{inl}(v)}}
\newcommand{\inrv}[1]{\ensuremath{\mathit{inr}(v)}}
\newcommand{\ket}[1]{|#1\rangle}
\newcommand{\oneover}[1]{1/#1}
\newcommand{\pointed}[2]{[#1 \bullet #2]}
\newcommand{\nboxtimes}[2]{\,\,~{^{#1}\boxtimes^{#2}}~\,\,}
\newcommand{\mm}{\texttt{\textminus}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\inl}[1]{\textsf{inl}~#1}
\newcommand{\inr}[1]{\textsf{inr}~#1}
\newcommand{\idv}[3]{#2 \xrightarrow{#1} #3}
\newcommand{\cp}[3]{#1\stackrel{#2}{\bullet}#3}
\newcommand{\idt}[3]{#2 \equiv_{#1} #3}
\newcommand{\idrt}[3]{#3 \equiv_{#1} #2}
\newcommand{\refl}[1]{\textsf{refl}~#1}
\newcommand{\lid}{\textsf{lid}}
\newcommand{\rid}{\textsf{rid}}
\newcommand{\linv}{l!}
\newcommand{\rinv}{r!}
\newcommand{\invinv}{!!}
\newcommand{\assoc}{\circ}
\newcommand{\identlp}{\mathit{unite}_+\mathit{l}}
\newcommand{\identrp}{\mathit{uniti}_+\mathit{l}}
\newcommand{\identlsp}{\mathit{unite}_+\mathit{r}}
\newcommand{\identrsp}{\mathit{uniti}_+\mathit{r}}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{unite}_*\mathit{l}}
\newcommand{\identrt}{\mathit{uniti}_*\mathit{l}}
\newcommand{\identlst}{\mathit{unite}_*\mathit{r}}
\newcommand{\identrst}{\mathit{uniti}_*\mathit{r}}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\absorbr}{\mathit{absorbr}}
\newcommand{\absorbl}{\mathit{absorbl}}
\newcommand{\factorzr}{\mathit{factorzr}}
\newcommand{\factorzl}{\mathit{factorzl}}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\distl}{\mathit{distl}}
\newcommand{\factorl}{\mathit{factorl}}
\newcommand{\distz}{\mathit{absorbr}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\proves}{\vdash}
\newcommand{\idc}{\mathit{id}\!\!\leftrightarrow}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2 \\\end{array}}
{\begin{array}{l}#3      \\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}

@@

 Uses of ancillae:
\url{https://quantum.country/search}:

\paragraph*{Reversible Branching Operations.}

\begin{quote}
  There’s a rough heuristic worth noting here, which is that you can
  often convert if-then style thinking into quantum circuits. You
  introduce an ancilla qubit to store the outcome of evaluating the if
  condition. And then depending on the state of the ancilla, you
  perform the appropriate state manipulation. Finally, when possible
  you reverse the initial computation, resetting the ancilla to its
  original state so you can subsequently ignore it.
\end{quote}

\paragraph*{Error Correction.}

\url{https://www.nap.edu/read/25196/chapter/5#73}
\begin{quote}
  For error correction, one needs to replicate the state of a qubit
  onto a number of qubits. While the no cloning theorem prevents one
  from copying the state of a qubit directly onto another, one can
  create a redundant entangled qubit state of many qubits. The key is
  that the qubits to be entangled must start out in a known
  state. Qubits with a known state (for purposes of this discussion,
  it will be the state $\ket{0}$), called ``ancilla qubits,'' may be
  added to a computation for this purpose. Since the state of ancilla
  qubits are known, it is possible to create a simple circuit that
  makes the output state of all these ancilla qubits match the
  protected qubit: run each ancilla through a controlled-NOT gate,
  where the control is driven by the qubit that needs to be
  replicated. Assume that there is a qubit with state $\Psi$ that we
  want to protect, where $\ket{\Psi}$ represents an arbitrary
  superposition state $\ket{\Psi} = a_0 \ket{0} + a_1 \ket{1}$. In the
  CNOT gate, the ancilla $\ket{0}$ state will remain a $\ket{0}$ state
  by the $\ket{0}$ component of $\ket{\Psi}$, but it will be converted
  to $\ket{1}$ by the $\ket{1}$ component of $\ket{\Psi}$. The result
  of this operation is the newly entangled two-qubit state
  $a_0 \ket{0} + a_1 \ket{1}$, creating a system in which the ancilla
  qubit is now perfectly entangled with the first qubit. Adding more
  ancillae increases the distance of the repetition code.  \end{quote}

\url{https://www.sigarch.org/the-case-for-quantum-computing/}
\begin{quote}
  Imagine a 3-qubit quantum majority code in which a logical “0” is
  encoded as “000” and a logical “1” is encoded as “111.”  Just as
  with a classical majority code, a single bit-flip error can be
  corrected by restoring to the majority value.  Unlike a classical
  code, however, we can not directly measure the qubits else their
  quantum state will be destroyed.  Instead, we measure syndromes from
  each possible pair of qubits by interacting them with an ancilla,
  then measure each ancilla.  Although the errors to the qubits are
  actually continuous, the effect of measuring the ancilla is to
  discretize the errors, as well as inform us whether an error
  occurred so that it can be corrected.  With this methodology,
  quantum states are restored in a modular way for even a large
  quantum computer.  Furthermore, operations on error-corrected qubits
  can be viewed as digital rather than analog, and only a small number
  of universal operations (H, T, CNOT) are needed for universal
  quantum computation.  Through careful design and engineering, error
  correction codes and this small set of precise operations will lead
  to machines that could support practical quantum computation.
\end{quote}

\paragraph*{Deferring Quantum Measurements.}

\url{https://homepages.cwi.nl/~rdewolf/qcnotesv2.pdf}
\begin{quote}
  Using an ancilla qubit, it is possible to avoid doing any
  intermediate measurements in a quantum circuit. Show how this can be
  done.  Hint: instead of measuring the qubit, apply a CNOT that
  “copies” it to a new $\ket{0}$-qubit, which is then left alone until
  the end of the computation. Analyze what happens.
\end{quote}

%%%
\subsection{Language Support}

Leads to bugs as analyzed by
\url{http://drops.dagstuhl.de/opus/volltexte/2019/10196/pdf/OASIcs-PLATEAU-2018-4.pdf}:

\begin{quote}

  Bug type 6: Incorrect composition of operations using mirroring Section 4.5 discussed how bugs in deallocating ancillary qubits can happen due to bad parameters. Here we see how bugs in deallocating ancillary qubits can happen due to incorrect composition of operations following a mirroring pattern. For example, in Table 7, the operations in rows 2 and 3 are respectively mirrored and undone in rows 6 and 5. These lines of code need careful reversal of every loop and every operation.

  A common pattern in quantum programs involves performing operations (e.g., add), contingent on a set of qubits known as control qubits. Without language support, this pattern needs many lines of code and manual allocation of ancillary qubits. In the Scaffold code example in Table 7, rows 3 and 5 are just computing the intersection of qubits q, with the help of ancillary qubits initialized in row 1, in order to realize the controlled rotation operation in row 4. Furthermore, quantum algorithms often need varying numbers of control qubits in different parts of the algorithm, leading to replicated code from multiple versions of the same subroutine differing only by the number of control qubits.

\end{quote}

>>>>

\newcommand{\alt}{~|~}
\newcommand{\net}{t_\bullet}
\newcommand{\opr}[1]{|#1\rangle\langle#1|}
\newcommand{\inlv}[1]{\ensuremath{\mathit{inl}(v)}}
\newcommand{\inrv}[1]{\ensuremath{\mathit{inr}(v)}}
\newcommand{\ket}[1]{|#1\rangle}
\newcommand{\oneover}[1]{1/#1}
\newcommand{\pointed}[2]{[#1 \bullet #2]}
\newcommand{\nboxtimes}[2]{\,\,~{^{#1}\boxtimes^{#2}}~\,\,}
\newcommand{\mm}{\texttt{\textminus}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\inl}[1]{\textsf{inl}~#1}
\newcommand{\inr}[1]{\textsf{inr}~#1}
\newcommand{\idv}[3]{#2 \xrightarrow{#1} #3}
\newcommand{\cp}[3]{#1\stackrel{#2}{\bullet}#3}
\newcommand{\idt}[3]{#2 \equiv_{#1} #3}
\newcommand{\idrt}[3]{#3 \equiv_{#1} #2}
\newcommand{\refl}[1]{\textsf{refl}~#1}
\newcommand{\lid}{\textsf{lid}}
\newcommand{\rid}{\textsf{rid}}
\newcommand{\linv}{l!}
\newcommand{\rinv}{r!}
\newcommand{\invinv}{!!}
\newcommand{\assoc}{\circ}
\newcommand{\identlp}{\mathit{unite}_+\mathit{l}}
\newcommand{\identrp}{\mathit{uniti}_+\mathit{l}}
\newcommand{\identlsp}{\mathit{unite}_+\mathit{r}}
\newcommand{\identrsp}{\mathit{uniti}_+\mathit{r}}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{unite}_*\mathit{l}}
\newcommand{\identrt}{\mathit{uniti}_*\mathit{l}}
\newcommand{\identlst}{\mathit{unite}_*\mathit{r}}
\newcommand{\identrst}{\mathit{uniti}_*\mathit{r}}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\absorbr}{\mathit{absorbr}}
\newcommand{\absorbl}{\mathit{absorbl}}
\newcommand{\factorzr}{\mathit{factorzr}}
\newcommand{\factorzl}{\mathit{factorzl}}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\distl}{\mathit{distl}}
\newcommand{\factorl}{\mathit{factorl}}
\newcommand{\distz}{\mathit{absorbr}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\proves}{\vdash}
\newcommand{\idc}{\mathit{id}\!\!\leftrightarrow}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2 \\\end{array}}
{\begin{array}{l}#3      \\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}

@@

\paragraph*{Quantum blackboxes and Phase Kickback Trick.} Consider a
small database with four elements $a$, $b$, $c$, and $d$. We are given
a function $f$ that maps every element to $0$ except for one element
of interest, e.g., $f(a)=1$ and $f(b)=f(c)=f(d)=0$. In the worst case,
finding the element of interest might require applying $f$ to every
element of the database. In the quantum world we can find the element
much more efficiently using Grover's
algorithm~\cite{Grover:1996:FQM:237814.237866}. We start by building
an equally weighted superposition of the elements:
$\frac{1}{2}\ket{a}+\frac{1}{2}\ket{b}+\frac{1}{2}\ket{c}+\frac{1}{2}\ket{d}$
and operate concurrently of the superposition as follows:

\[\begin{array}{l}
                    \frac{1}{2}\ket{a}+\frac{1}{2}\ket{b}+\frac{1}{2}\ket{c}+\frac{1}{2}\ket{d}\\
\\
\Rightarrow  \textrm{phase kickback} \\
\\
-\frac{1}{2}\ket{a}+\frac{1}{2}\ket{b}+\frac{1}{2}\ket{c}+\frac{1}{2}\ket{d}\\
\\
\Rightarrow  \textrm{diffusion} \\
\\
 (\frac{1}{2}-(-\frac{1}{2}))\ket{a}+(\frac{1}{2}-\frac{1}{2})\ket{b}+(\frac{1}{2}-\frac{1}{2})\ket{c}+(\frac{1}{2}-\frac{1}{2})\ket{d})\\
\\
\Rightarrow  \textrm{simplification} \\
\\
\ket{a}
\end{array}\]

The question, which arises repeatedly in quantum computation, is how
to implement operations such as phase kickback. Essentially we want to
concurrently perform the following action on every element of the
superposition:
\[\begin{array}{l}
  \textbf{if}~\textit{current-element}=\ket{a} \\
  \textbf{then}~\textit{negate} \\
  \textbf{else}~\textit{identity}
\end{array}\]
In the quantum world, na\"\i vely attempting to test whether the
current element is equal to $\ket{a}$ by performing a measurement would
destroy the quantum superposition defeating the entire algorithm.

Here is where ancilla qubits come to the rescue. In the words of
Matuschak and Nielsen~\cite{howgrover}, the idea is the following:
\begin{quote}
  There's a rough heuristic worth noting here, which is that you can
  often convert \verb|if-then| style of thinking into quantum
  circuits. You introduce an ancilla qubit to store the outcome of
  evaluating the \verb|if| condition. And then depending on the state
  of the ancilla, you perform the appropriate state
  manipulation. Finally, when possible you reverse the initial
  computation, resetting the ancilla to its original state so you can
  subsequently ignore it.
\end{quote}

\paragraph*{Quantum Error Correction.} The main requirement for
quantum error correction is the ability to replicate the state of a
qubit onto a number of qubits~\cite[Ch.~3]{NAP25196}. This is however
impossible to do directly by the no-cloning theorem. Again ancilla
qubits come to the rescue: because ancilla qubits start in a known
initial state, it is possible to create a simple circuit that makes
their output state match a protected qubit without knowing or
disturbing the protected qubit. Chong et al.~\cite{sigarchblog}
illustrate the idea with a simple example:
\begin{quote}
consider a 3-qubit quantum majority code in which a logical ``0'' is
encoded as ``000'' and a logical ``1'' is encoded as ``111.''  Just as with
a classical majority code, a single bit-flip error can be corrected by
restoring to the majority value.  Unlike a classical code, however, we
can not directly measure the qubits else their quantum state will be
destroyed.  Instead, we measure syndromes from each possible pair of
qubits by interacting them with an ancilla, then measure each ancilla.
Although the errors to the qubits are actually continuous, the effect
of measuring the ancilla is to discretize the errors, as well as
inform us whether an error occurred so that it can be corrected.  
\end{quote}

\paragraph*{Deferring Quantum Measurements.} The evolution of a
quantum system is deterministic. Measurement add significant
complications, collapsing the quantum state, and producing
probabilistic results. However, using ancilla qubits it is possible to
avoid doing any intermediate measurements in a quantum
circuit~\cite{dewolf}.

%%%
\subsection{Ancillae Bugs in Benchmarks.}
Despite its fundamental importance, the management of ancillae is
poorly supported in current languages, simulators, and toolkits as
demonstrated in a recent analysis of bugs in quantum
benchmarks~\cite{DBLP:conf/oopsla/HuangM18}. The analysis reveals that
lack of language support in the management of ancillae leads to
several problems. For example: 
\begin{quote}
  \textbf{Bug type 6: Incorrect composition of operations using mirroring}
  Section~4.5 discussed how bugs in deallocating ancillary qubits can
  happen due to bad parameters. Here we see how bugs in deallocating
  ancillary qubits can happen due to incorrect composition of
  operations following a mirroring pattern. For example, in Table~7,
  the operations in rows 2 and 3 are respectively mirrored and undone
  in rows~6 and~5. These lines of code need careful reversal of every
  loop and every operation.
\end{quote}
